The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-03-19 src/ms.tex
Address #53 on paragraph apy on units
Signed-off-by: nstarman <nstarkman@protonmail.com>

\begin{abstract}
The \astropy Project supports and fosters the development of open-source and openly developed
\python packages that provide commonly needed functionality to the astronomical
community.
A key element of the \astropy Project is the core package \astropypkg, which serves as the
foundation for more specialized projects and packages.
In this article, we summarize key features in the core package as of the recent major
release, version 5.0, and provide major updates for the project.
We then discuss supporting a broader ecosystem of interoperable packages,
including connections with several astronomical observatories and missions.
We also revisit the future outlook of the \astropy Project and the current
status of Learn Astropy.
We conclude by raising and discussing the current and future challenges facing the project.
\end{abstract}

@@

The \python programming language is a high-level, interpreted (as opposed to
compiled) programming language that has become an industry standard across many
computational domains, technological sectors, and fields of research.
This recent and rapid adoption of \python stems from the fact that it enables
scalable, time- and energy-efficient code execution \citep[e.g.,][]{Augier:2021}
with a focus on code readability, ease of use, and interoperability with other
languages.
Over the last decade, \python has grown enormously in popularity to become a
dominant programming language, especially in the astronomical and broader scientific
communities.
For example, Figure~\ref{fig:python-mentions} shows the number of yearly
full-text mentions of \python as compared to a selection of other programming languages
in refereed articles in the astronomical literature, demonstrating its nearly
exponential growth in popularity.
The adoption of \python by astronomy researchers, students, observatories, and
technical staff combined with an associated increase in awareness and interest
about open-source software tools is contributing to a paradigm shift in the way
research is done, data is analyzed, and results are shared in astronomy and
beyond.

One of the factors that has led to its rapid ascent in popularity in scientific
contexts has been the significant, volunteer-driven effort behind developing
community-oriented open-source software tools and fostering communities of users
and developers that have grown around these efforts.
Today, a broad and feature-diverse ``ecosystem'' of packages exists in the
\python scientific computing landscape: roughly ordered from general-use to
domain-specific, this landscape now includes packages that provide core
numerical analysis functionality like \package{numpy} \citep{numpy:nature} and
\package{scipy} \citep{scipy}, visualization frameworks like
\package{matplotlib} \citep{matplotlib}, machine learning and data analysis
packages like \package{tensorflow} \citep{tensorflow}, \package{pymc3}
\citep{Salvatier:2016}, and \package{emcee} \citep{emcee}, domain-specific
libraries like \package{yt} \citep{yt:2011}, \package{plasmapy}
\citep{plasmapy}, \package{sunpy} \citep{sunpy:apj}, \package{Biopython}
\citep{biopython}, and \package{sympy} \citep{sympy} (to name a few in each
category).
The \astropypkg \citep{astropy:2013, astropy:2018} core package began in this
vein, as an effort to consolidate the development of commonly used functionality
needed to perform astronomical research into a community-developed \python
package.

The \astropypkg core package was one of the first large, open-source \python
packages developed for astronomy and provides, among other things, software
functionality for reading and writing astronomy-specific data formats (e.g.,
FITS), transforming and representing astronomical coordinates, and representing
and propagating physical units in code.
An early description of the core functionality in \astropypkg can be found in
the first \astropy paper \citep{astropy:2013} or in detail in the core package
documentation.\footnote{\url{https://docs.astropy.org/}}
The codebase in the \astropypkg core package is now largely stable, in that the
software interface does not change without sufficient and significant
motivation, and the addition of new features into the core package has slowed
as compared to the first years of its development.
This is largely driven by the fact that the core package now represents just one
piece of the broader astronomy \python context, and new feature development
is now happening in more specialized packages that are expanding the
capabilities of the \astropy ecosystem by building on top of the foundations
laid by the \astropypkg core package.
Because of this natural expansion, the name \astropy has grown in scope beyond a
single \python library to become ``the Astropy Project.''

@@

The \astropy Project is a community effort that represents the union of the
\astropypkg core package, the ecosystem of astronomy-specific software tools
that are interoperable with \astropypkg (Astropy Coordinated and Affiliated Packages),
\emph{and} the community of users, developers, and maintainers that participate
in Astropy efforts.
However, there is no single institution responsible for managing the Astropy Project,
for funding or maintaining its development, or sustaining it into the future:
The Project is maintained and coordinated largely by volunteers and contributors
who are primarily paid for other reasons (i.e., Astropy work is just a means to an end).
While new Astropy-affiliated packages are being developed that expand upon the
core functionality in the \astropypkg package, representing a natural expansion
of the Astropy Project ecosystem, the needs of and challenges faced by the
Project are evolving.
In particular, the transition from focusing our energy on development and
maintenance of a single core package, to instead sustaining the core package and
fostering the development of the community and its expansion has been a key
issue faced by the Astropy Project in the last several years.

@@

Major versions of the core package --- that is, versions that add and/or modify
functionality --- are released approximately every six months, and are then
maintained with releases that fix issues until the next major version is
released. However, every two years a major release is designated as a long-term
support (LTS) release, which continues to be maintained for up to two years
\citep{ape2}. The motivation for LTS releases is to provide longer-term stable
versions of \astropypkg that users who require a high level of stability can
make use of if they do not always need the latest features (this includes but
not limited to telescope operations and data reduction pipelines). The v5.0
release of the core package, replacing v4.0, was designated as LTS at the end of 2021;
it will be maintained until the end of 2023.

@@

The \astropypkg core package is mature and stable, in that many features have
been part of the core package for years with purposefully few changes to the
software interface. This maturity and stability allows for the broader
astronomy \python community to rely upon the \astropypkg core package and build
specialized packages within the Astropy ecosystem. Even many non-astronomical
\python libraries have come to rely on \astropypkg. The relationship between
\astropy and the broader \python community is reciprocal, with the evolving needs
of the community and the maturation of the broader scientific \python ecosystem
driving much of the development of the \astropypkg core package.

Within \astropypkg, development may be roughly split into several
categories: new features, intra- and inter-package interoperability,
improvements to precision and reproducibility, performance and documentation
enhancements, and bug fixes. We discuss some of these in the subsections below.

Additionally, some features have been moved from the core package, either
because they are of more general use outside of \astropypkg, or because they are
too specialized and belong in an Astropy-affiliated package. An example of the
former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA) software
Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that \astropy
carried, as well as the associated \python wrappers. Since this is basic
infrastructure, with a release cycle set by SOFA, it made more sense to create
separate packages that serve as \astropypkg dependencies: \package{ERFA}
\citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
was too specialized was a Virtual Observatory sub-package; the Simple
Application Message Protocol (\astropysubpkg{samp}) is kept in the core package
for now, but other parts more properly belonged in
\package{astroquery} (see Section~\ref{sec:astroquery}).

@@

New features present a particular difficulty for a mature package such as
\astropypkg, because it is challenging to know {\em a priori} what the best
interface will be, and hard to address all possible use cases in one round of
development, even within \astropypkg\ itself. New features come in various forms,
from new submodules providing wholly new capabilities, to larger additions to
existing submodules, to large rewrites of the interface of modules. This range
is spanned by five main new features that entered the \astropypkg core since the
previous summary \citepalias{astropy:2018}. In our descriptions of each, we
include how we hope the new feature will evolve. Ultimately, in a user-driven
project like \astropy, development depends primarily on the priorities of
contributors.

\paragraph{Uncertainties and Distributions} In scientific analysis,
measurement values with units and uncertainties are essential.
\astropyunits provides the ability to associate numbers with units, and
propagate these correctly, using the \astropyQuantity\ class. However, the
ability to associate and propagate uncertainties is relatively limited: in
\astropysubpkg{nddata}, there are options to associate errors with data arrays,
but covariances are not tracked.

@@

A plan for future development of \astropysubpkg{uncertainties} is to
support error propagation: tracking covariances for the case that uncertainties
are normally distributed rather than approximated by Monte-Carlo means. One
implementation problem to be solved is when to {\em stop} tracking covariances.
For instance, a simple operation such as subtracting the mean from $N$ data
points implies that all data points are now covariant with each other, i.e., one
has to carry $N\times N$ covariances. For a large image, that becomes unnecessary
and computationally expensive.

\paragraph{Masked quantities}  It can be useful to mask bad data, either by marking
bad data values with a flag or by replacing them with a special
value, e.g., ``Not a Number'' (NaN). In \astropypkg, both approaches are
used: a flag for masked columns in \astropyTable, mask flags and bitmaps
for N-dimensional data in \astropysubpkg{nddata}, and replacing elements with
NaN in \astropyTime. However masked \astropyQuantity support is currently very limited,
hindering usage of masks in a \astropyQTable\ (in which \astropyQuantity\ is used
for all columns with units).

Furthermore, the \texttt{MaskedArray} class from \package{numpy} only works
well with plain data arrays, hiding other metadata and attributes, for example
the \texttt{unit} of a \astropyQuantity. Consequently, a new \astropyMasked\
class was designed, based on a framework very similar to that of
\astropyDistribution, making it easy to create masked instances of other
classes. In particular, \astropyMasked\ can be used to create masked quantities
for \astropyQTable, enabling I/O of masked data from different file types.

The masked quantities work largely without any changes in higher-level objects
such as \astropySkyCoord, but more work is needed to better integrate it with
the rest of \astropypkg, e.g., using masked arrays in \astropyTime\ instead of NaN.

\paragraph{Time series}  From sampling a continuous variable at fixed times to
counting events binned into time windows, many different areas of astrophysics
require the manipulation of time series. The new \astropysubpkg{timeseries}
sub-package extends the \astropyQTable\ class to support tables of data as a
function of time, where the data can either represent samples or averages over
particular time bins. The new classes offer a number of special methods to
manipulate time series (e.g., folding and resampling) and to read different data
formats (e.g., \mission{Kepler} light curves).

@@

\paragraph{A High-Level Interface to World Coordinate Systems}
Astronomical data are often provided alongside information about the
correspondence between ``real-world'' and pixel coordinates. This mapping is the
essence of the World Coordinate System (WCS) concept. From its inception,
the \astropysubpkg{wcs} sub-package allowed access to WCS information provided
in, e.g., FITS files, but it became clear that other WCS standards and
representations had to be supported for new missions and observatories (e.g.,
the James Webb Space Telescope and the Rubin Observatory). To harmonize these
needs, a new high-level interface was created based on a formal design first
proposed in an \astropy\ Proposal for Enhancement (\astropyAPE{14}). The hope is
that by having a formal design, other packages implementing WCS objects can
straightforwardly modify their classes to conform to the new interface or build
thin wrappers that conform. The implementation in \astropysubpkg{wcs} interacts
well with other \astropypkg\ objects such as \astropySkyCoord\ and \astropyTime.
Future work includes better integration of this interface with the rest of
the Project.

@@

Because \astropy is modular and situated at the nexus between scientific
computing and astronomy, interoperability is an important focus for \astropy,
both within the various \astropypkg sub-packages and with other \python
libraries. We aim for a seamless user experience, where different code blocks
would ``just work'' when put together. Here, we describe a few particular
efforts towards this goal.

@@

\astropyQuantity\ is a backbone of \astropy, leveraging the power of
\package{numpy} and adding units. Previously, many \package{numpy} functions
would strip a \astropyQuantity\ of its units (or fail outright), limiting
\astropyQuantity's potential. Now, advances in \package{numpy} function
overloading (e.g.,
\href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
mean \astropyQuantity\ works with almost all \package{numpy} (v1.17+)
functions.

@@

Units were also integrated further within \astropypkg. For an already-defined
unit-less \astropyModel, e.g., those imported from another library,
\astropymodeling can now coerce units. A new module – \astropycosmologyunits –
has been added to the cosmology sub-package for defining and collecting
cosmological units and equivalencies. \astropycosmologyunits has a unit and
equivalencies for \texttt{littleh}, and a new unit, \texttt{redshift}, for
factors of cosmological redshift. \texttt{redshift} has a number of
equivalencies for converting between cosmological distance measures, e.g., CMB
temperature or comoving distance.

@@

Within \astropypkg and \python, there are numerous ways to represent and store
array-valued data. Some of the differences are historical:
\astropyapidoc{table.Column}{table.Column} and
\astropyapidoc{io.fits.Column}{io.fits.Column} are not the same. Some
differences are computational: \package{numpy}, \package{cupy}, and
\package{dask} arrays have almost identical APIs, but are optimized for
different use cases. Lastly, some differences are inherent: \astropyQuantity,
\astropyTime, and \astropySkyCoord\ represent fundamentally different types of
objects.

We aim to make it possible for any array-valued data to be used as a column in a
table. A well defined protocol for mixin-columns has been developed for
\astropytable, allowing the original object to be used as a column with a
familiar API, and to ``round-trip'' through tables with no loss of data or
attributes. With this protocol, it is now possible to store \astropy native
objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord\ --
within a \astropyTable\ and write these to various file formats, such as FITS.
For objects not already covered by the mixin protocol, functions can be
registered with \astropyTable\ to convert any array-like object into a mixin
column. As an example, the mixin functions are used to integrate \package{dask}
arrays with \astropyTable, allowing cloud-stored or cluster-scale data to be
used as a column in a \astropyTable.

@@

The FITS standard was extended to rigorously describe time coordinates in the
WCS framework \citep{FITS-Time:2015}. Compared to
other types of coordinates in WCS, time requires more metadata: format, scale,
position, reference, etc. This metadata had to be manually specified and the
nuances understood by the user. \astropyfits could read this data as a standard
\astropyFitsColumn\, but would not interpret the time-related metadata and
attributes. Through the support of the
\href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
Summer of Code} 2017 program,\footnote{ To learn more about this project, please
see the final report
\href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
mixin protocol for seamless interoperability}.} the \astropyfits package now
interprets time data correctly, using \astropyTime\ as a mixin column. For
backward compatibility, this feature may be turned off.

@@

\astropy now supports reading and writing tables in the American Astronomical
Society Journals' Machine-Readable Table (\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}) format. This ASCII format has
long been missing and was added to \astropyTable\ I/O in a Google Summer of Code 2021 project.
In addition, \astropyTable\ may also read from and write
to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
\href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},
and \href{https://wwwastro.msfc.nasa.gov/qdp/}{QDP} formats.

The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
standard} has been updated to v1.0, adding support of three additional data
subtypes: multidimensional column data (both masked and unmasked)
with fixed dimensions in all table cells, multidimensional
column data with variable-dimension arrays similar to FITS variable-length
arrays, and object-type columns with simple \python objects.

As mentioned previously, \package{dask} arrays may be used as a mixin column in
\astropyTable. Now, \package{dask} may also be a data array in
\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and only
computed while written, avoiding excessive memory use. Furthermore, \astropyTable\ can
be appended to an existing FITS file, where \package{dask} mixin columns can
interoperate seamlessly between them.

@@

\astropyiosubpkg{registry} is a powerful way to define input and output (I/O)
functions for \astropypkg objects, such as reading from or writing to a file,
following a given standard. \astropypkg uses this internally for the I/O methods in
\astropyTable\ and \astropyCosmology. Users can also register custom I/O to
extend the options on these classes.

As of \astropypkg v5.0, the I/O registry submodule has been generalized to enable a
number of new use cases; for instance, the class-based architecture allows
streamlining of the creation of custom registries while minimizing code duplication.
One application is in \astropycosmology,
which has two different \textit{kinds} of registries for \astropyCosmology: one
for reading and writing files, and another for converting between \python objects.

\subsection{Precision \& Reproducibility}
\label{sec:core-features-precision_accuracy_repreducibility}

@@

\paragraph{Time}
In astronomy, time accuracy down to the (nano)second is frequently important.
Whether for planning observations or crunching pulsar or VLBI data, missing
seconds meaningfully impact results. In past versions of \astropypkg, when using
\astropyTime\, leap seconds had to be manually applied; now, leap
seconds are applied automatically. Moreover, \astropypkg updates internal time
data files to ensure that the correct leap second adjustment is always used. For
reproducibility with older code, \astropyLeapSeconds\ may also be
manually applied.

@@

Measurements of physical constants (and the units defined from them) improve
over time. Periodically the standardized systems of units and constants are
updated to reflect these improvements. For instance, in 2019 the SI system was
redefined \citep{NIST2019}, with an accompanying update to the physical
constants in
\href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
2018}}. \astropyconstants now defaults to using the \texttt{SI/CODATA 2018}
values, with the units in \astropyunits based on these constants. Most of
\astropypkg and affiliate packages build upon \astropyunits, so this update
affects the entire \astropy ecosystem.

@@

Performance in \astropypkg is addressed based on feedback and requests, and
through contributions from community members: if you find that parts of
\astropypkg are not performant, we encourage you to open a \github issue to
start a discussion with the \astropypkg core maintainers.

@@

The Astropy project supports the broader ecosystem by providing pre-configured
infrastructure packages that the community can use to support and maintain their
own software package infrastructure. These include tools and extensions that
enable generating documentation and setting up automated testing, as well
as providing package scaffolding for new projects.

@@

We would like to thank the members of the community who have contributed to
\astropy, opened issues and provided feedback, and supported the
project in a number of different ways.

>>>>

\begin{abstract}
The \astropy Project supports and fosters the development of open-source and openly-developed
\python packages that provide commonly-needed functionality to the astronomical
community.
A key element of the \astropy Project is the core package \astropypkg, which serves as the
foundation for more specialized projects and packages.
In this article, we summarize key features in the core package as of the recent major
release, version 5.0, and provide major updates for the project.
We then discuss supporting a broader ecosystem of inter-operable packages,
including connections with several astronomical observatories and missions.
We also revisit the future outlook of the \astropy Project and the current
status of Learn Astropy.
We conclude by raising and discussing the current and future challenges facing the project.
\end{abstract}

@@

The \python programming language is a high-level, interpreted (as opposed to
compiled) programming language that has become an industry standard across many
computational domains, technological sectors, and fields of research.
This recent and rapid adoption of \python stems from the fact that it enables
scalable, time- and energy-efficient code execution \citep[e.g.,][]{Augier:2021}
with a focus on code readability, ease of use, and interoperability with other
languages.
Over the last decade, \python has grown enormously in popularity to become a
dominant programming language also in the astronomical and broader scientific
communities.
For example, Figure~\ref{fig:python-mentions} shows the number of yearly
full-text mentions of \python as compared to a few other programming languages
in refereed articles in the astronomical literature, demonstrating its nearly
exponential growth in popularity.
The adoption of \python by astronomy researchers, students, observatories, and
technical staff combined with an associated increase in awareness and interest
about open-source software tools is contributing to a paradigm shift in the way
research is done, data is analyzed, and results are shared in astronomy and
beyond.

One of the factors that has led to its rapid ascent in popularity in scientific
contexts has been the significant, volunteer-driven effort behind developing
community-oriented open-source software tools and fostering communities of users
and developers that have grown around these efforts.
Today, a broad and feature-diverse ``ecosystem'' of packages exists in the
\python scientific computing landscape: Roughly ordered from general-use to
domain-specific, this landscape now includes packages that provide core
numerical analysis functionality like \package{numpy} \citep{numpy:nature} and
\package{scipy} \citep{scipy}, visualization frameworks like
\package{matplotlib} \citep{matplotlib}, machine learning and data analysis
packages like \package{tensorflow} \citep{tensorflow}, \package{pymc3}
\citep{Salvatier:2016}, and \package{emcee} \citep{emcee}, domain-specific
libraries like \package{yt} \citep{yt:2011}, \package{plasmapy}
\citep{plasmapy}, \package{sunpy} \citep{sunpy:apj}, \package{Biopython}
\citep{biopython}, and \package{sympy} \citep{sympy} (to name a few in each
category).
The \astropypkg \citep{astropy:2013, astropy:2018} core package began in this
vein, as an effort to consolidate the development of commonly-used functionality
needed to perform astronomical research into a community-developed \python
package.

The \astropypkg core package was one of the first large, open-source \python
packages developed for astronomy and provides, among other things, software
functionality for reading and writing astronomy-specific data formats (e.g.,
FITS), transforming and representing astronomical coordinates, and representing
and propagating physical units in code.
An early description of the core functionality in \astropypkg can be found in
the first \astropy paper \citep{astropy:2013} or in detail in the core package
documentation.\footnote{\url{https://docs.astropy.org/}}
The codebase in the \astropypkg core package is now largely stable, in that the
software interface does not change without sufficient and significant
motivation, and the addition of new features into the core package has slowed
as compared to the first years of its development.
This is largely driven by the fact that the core package now represents just one
piece of the broader astronomy \python context, and much new feature development
is now happening in more specialized packages that are expanding the
capabilities of the \astropy ecosystem by building on top of the foundations
laid by the \astropypkg core package.
Because of this natural expansion, the name \astropy has grown in scope beyond a
single \python library to become ``the Astropy Project.''

@@

The \astropy Project is a community effort that represents the union of the
\astropypkg core package, the ecosystem of astronomy-specific software tools
that are interoperable with \astropypkg (Astropy Affiliated Packages),
\emph{and} the community of users, developers, and maintainers that participate
in Astropy efforts.
However, there is no institution responsible for managing the Astropy Project,
for funding or maintaining its development, or sustaining it into the future:
The Project is maintained and coordinated largely by volunteers.
While new Astropy-affiliated packages are being developed that expand upon the
core functionality in the \astropypkg package, representing a natural expansion
of the Astropy Project ecosystem, the needs of and challenges faced by the
Project are evolving.
In particular, the transition from focusing our energy on development and
maintenance of a single core package, to instead sustaining the core package and
fostering the development of the community and its expansion has been a key
issue faced by the Astropy Project in the last several years.

@@

Major versions of the core package --- that is, versions that add and/or modify
functionality --- are released approximately every six months, and are then
maintained with releases that fix issues until the next major version is
released. However, every two years a major release is designated as a long-term
support (LTS) release, which continues to be maintained for up to two years
\citep{ape2}. The motivation for LTS releases is to provide longer-term stable
versions of \astropypkg that users that require a high level of stability can
make use of if they do not always need the latest features (this could include,
for example, telescope operation or data reduction pipelines, etc.). The v5.0
release of the core package was designated as LTS, and since it was released at
the end of 2021, it will be maintained until the end of 2023.

@@

The \astropypkg core package is mature and stable, in that many features have
been part of the core package for years with purposefully few changes to the
software interface. This maturity and stability allows for the broader
astronomy \python community to rely upon the \astropypkg core package and build
specialized packages within the Astropy ecosystem. Even many non-astronomical
\python libraries have come to rely on \astropypkg. The relationship between
\astropy and the \python-using community is reciprocal, with the evolving needs
of the community and the maturation of the broader scientific \python ecosystem
driving much of the development of the \astropypkg core package.

Within \astropypkg, development may be roughly split into a few
categories: new features added, intra- and inter-package interoperability,
improvements to precision, accuracy, and reproducibility, and performance
enhancements. We discuss these in the subsections below.

Additionally, some features have been moved from the core package, either
because they are of more general use outside of \astropypkg, or because they are
too specialized and belong in an Astropy-affiliated package. An example of the
former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA) software
Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that \astropy
carried, as well as the \python wrappers for it. Since this is basic
infrastructure, with a release cycle set by SOFA, it made more sense to create
separate packages that serve as \astropypkg dependencies: \package{ERFA}
\citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
was too specialized was a Virtual Observatory sub-package; the Simple
Application Message Protocol was kept in the core package
(\astropysubpkg{samp}), but other parts more properly belonged in
\package{astroquery} (see Section~\ref{sec:astroquery}).

@@

New features present a particular difficulty for a mature package such as
\astropypkg, because it is difficult to know {\em a priori} what the best
interface will be, and hard to address all possible use cases in one round of
development, even within \astropypkg\ itself. New features come in various forms,
from new submodules providing wholly new capabilities, to larger additions to
existing submodules, to large rewrites of the interface of modules. This range
is spanned by five main new features that entered the \astropypkg core since the
previous summary \citepalias{astropy:2018}. In our descriptions of each we
include how we hope the new feature will evolve. Ultimately, in a user-driven
project like \astropy, development depends primarily on the priorities of
contributors.

\paragraph{Uncertainties and Distributions} Early in our careers we were likely
all taught that measurements without units and uncertainties are meaningless.
\astropyunits provides the ability to associate numbers with units, and
propagate these correctly, using the \astropyQuantity\ class. However, the
ability to associate and propagate uncertainties is relatively limited: in
\astropysubpkg{nddata}, there are options to associate errors with data arrays,
but these are treated as independent, i.e., covariances are not tracked.

@@

A plan for future development of \astropysubpkg{uncertainties} is to also
support error propagation, tracking covariances, for the case that uncertainties
are normally distributed rather than approximated by Monte-Carlo means. One
implementation problem to be solved is when to {\em stop} tracking covariances.
For instance, a simple operation such as subtracting the mean from $N$ data
points implies that all data points are now covariant with each other, i.e., one
has to carry $N\times N$ covariances. For a large image, that becomes not just
pointless but also prohibitive in terms of memory and CPU usage.

\paragraph{Masked quantities}  It can be useful to mask bad data. How one
supports masked data involves a number of choices, such as whether one indicates
data values are bad by a separate flag or by replacing bad data with a special
value, such as ``Not a Number'' (NaN). In \astropypkg, both approaches have
been used: a flag for masked columns in \astropyTable, mask flags and bitmaps
for N-dimensional data in \astropysubpkg{nddata}, and replacing elements with
NaN in \astropyTime. However masked quantities were poorly supported, making it
difficult to use masks in a \astropyQTable\ (in which \astropyQuantity\ is used
for all columns with units).

Unfortunately, the \texttt{MaskedArray} class from \package{numpy} only works
well with plain data arrays, hiding other metadata and attributes, for example
the \texttt{unit} of a \astropyQuantity. Consequently, a new \astropyMasked\
class was designed, based on a framework very similar to that of
\astropyDistribution, making it easy to create masked instances of other
classes. In particular, \astropyMasked\ can be used to create masked quantities
for \astropyQTable, enabling I/O of masked data from different file types.

The masked quantities work largely without any changes in higher-level objects
such as \astropySkyCoord, but work is still underway to expose the mask in those
classes, as well as to use masked arrays in \astropyTime\ instead of replacing
data with NaN.

\paragraph{Time series}  From sampling a continuous variable at fixed times to
counting events binned into time windows, many different areas of astrophysics
require the manipulation of time series. The new \astropysubpkg{timeseries}
sub-package extends the \astropyQTable\ class to support tables of data as a
function of time, where the data can either represent samples or averages over
particular time bins. The new classes offer a number of special methods to
manipulate time series (folding, resampling, etc.) and to read different data
formats (such as \mission{Kepler} light curves, etc.).

@@

\paragraph{A High-Level Interface to World Coordinate Systems}
Astronomical data are often provided alongside information about the
correspondence between ``real-world'' and pixel coordinates. This mapping is the
essence of the ``World Coordinate System (WCS)'' concept. From its inception,
the \astropysubpkg{wcs} sub-package allowed access to WCS information provided
in, e.g., FITS files, but it became clear that other WCS standards and
representations had to be supported for new missions and observatories (e.g.,
the James Webb Space Telescope and the Rubin Observatory). To harmonize these
needs, a new high-level interface was created based on a formal design first
proposed in an \astropy\ Proposal for Enhancement (\astropyAPE{14}). The hope is
that by having a formal design, other packages implementing WCS objects can
straightforwardly modify their classes to conform to the new interface or build
thin wrappers that conform. The implementation in \astropysubpkg{wcs} interacts
well with other \astropypkg\ objects such as \astropySkyCoord\ and \astropyTime.

@@

Because \astropy is modular and situated at the nexus between scientific
computing and astronomy, interoperability is an important focus for \astropy,
both within the various \astropypkg sub-packages and with other \python
libraries. If done right, all a person notices is that code using many different
python features and libraries ``just works.'' Here, we describe a few particular
efforts towards this goal.

@@

\astropyQuantity\ is a backbone of \astropy, leveraging the power of
\package{numpy} and adding units. Previously, many \package{numpy} functions
would strip a \astropyQuantity\ of its units (or fail outright), limiting
\astropyQuantity's usefulness. Now, advances in \package{numpy} function
overloading (e.g.,
\href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
mean \astropyQuantity\ works with almost all \package{numpy} (v1.17+)
functions.

@@

Units were also integrated further within \astropypkg. For an already-defined
unit-less \astropyModel, e.g. those imported from another library,
\astropymodeling can now coerce units. A new module – \astropycosmologyunits –
has been added to the cosmology sub-package for defining and collecting
cosmological units and equivalencies. \astropycosmologyunits has a unit and
equivalencies for \texttt{littleh}, and a new unit, \texttt{redshift}, for
factors of cosmological redshift. \texttt{redshift} has a number of
equivalencies for converting between cosmological distance measures, e.g. CMB
temperature or comoving distance.

@@

Within \astropypkg and \python there are numerous ways to represent and store
array-valued data. Some of the differences are historical:
\astropyapidoc{table.Column}{table.Column} and
\astropyapidoc{io.fits.Column}{io.fits.Column} are not the same. Some
differences are computational: \package{numpy}, \package{cupy}, and
\package{dask} arrays have almost identical APIs, but are optimized for
different use cases. Lastly, some differences are inherent: \astropyQuantity,
\astropyTime\, and \astropySkyCoord\ represent fundamentally different types of
objects.

We aim to make it possible for any array-valued data to be used as a column in a
table. A well defined protocol for mixin-columns has been developed for
\astropytable, allowing the original object to be used as a column with a
familiar API, and to ``round-trip'' through tables with no loss of data or
attributes. With this protocol, it is now possible to store \astropy native
objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord --
within a \astropyTable\ and write these to various file formats, such as FITS.
For objects not already covered by the mixin protocol, functions can be
registered with \astropyTable\ to convert any array-like object into a mixin
column. As an example, the mixin functions are used to integrate \package{dask}
arrays with \astropyTable, allowing cloud-stored or cluster-scale data to be
used as a column in a \astropyTable.

@@

The FITS standard was extended to rigorously describe time coordinates in the
World Coordinate System (WCS) framework \citep{FITS-Time:2015}. Compared to
other types of coordinates in WCS, time requires more metadata: format, scale,
position, reference, etc. This metadata had to be manually specified and the
nuances understood by the user. \astropyfits could read this data as a standard
\astropyFitsColumn\, but would not interpret the time-related metadata and
attributes. Through the support of the
\href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
Summer of Code} 2017 program,\footnote{ To learn more about this project, please
see the final report
\href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
mixin protocol for seamless interoperability}.}, the \astropyfits package now
interprets time data correctly, using \astropyTime\ as a `mixin' column. For
backward compatibility with manual systems, this feature may be turned off.

@@

\astropy now supports reading and writing tables in the American Astronomical
Society Journals' Machine-Readable Table (MRT) format. This ASCII format has
long been missing and as added in this year's Google Summer of Code project adds
\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}
to \astropyTable\ I/O. In addition, \astropyTable\ may also read from and write
to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
\href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},
and \href{https://wwwastro.msfc.nasa.gov/qdp/}{QDP} formats.

The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
standard} has been updated to version 1.0, adding support three additional data
subtypes. First, it adds multidimensional column data (both masked and unmasked)
with fixed dimensions in all table cells. Second, it adds multidimensional
column data with variable-dimension arrays similar to FITS variable-length
arrays. Third, it adds object-type columns with simple \python objects.

As mentioned previously, \package{dask} arrays may be used as a mixin column in
\astropyTable. Now, \package{dask} may also be the data array in
\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and if written to disk, the array
will be computed while written, avoiding excessive memory use. \astropyTable\ can
now be appended to an existing FITS file, and \package{dask} mixin columns
interoperate seamlessly between the two table types.

@@

\astropyiosubpkg{registry} is a powerful way to define input and output (I/O)
functions for \astropypkg objects, such as reading from or writing to a file,
following a given standard. \astropy uses this internally for the I/O methods in
\astropyTable\ and \astropyCosmology, and users can register custom I/O to
extend the options on these classes.

As of Astropy 5.0, the I/O registry submodule has been generalized to enable a
number of new use cases. for instance, the class-based architecture allows for
the creation of custom registries. One application is in \astropycosmology,
which has two different \textit{kinds} of registries for \astropyCosmology: one
for reading and writing files, and another for converting between python
objects. Using \astropyiosubpkg{registry} for custom classes means a developer
does not have to create a whole new I/O system and can offer to the user a
unified and familiar API.

\subsection{Precision, Accuracy, \& Reproducibility}
\label{sec:core-features-precision_accuracy_repreducibility}

@@

\paragraph{Time}
In astronomy, time accuracy down to the (nano)second is frequently important.
Whether for planning observations, crunching pulsar or VLBI data, missing
seconds meaningfully impact results. In past versions of \astropypkg, when using
\astropyTime\ in \astropytime leap seconds had to be manually applied; now, leap
seconds are applied automatically. Moreover, \astropypkg updates internal time
data-files to ensure that the correct leap second adjustment is always used. For
reproducibility, e.g., with an old code, \astropyLeapSeconds\ may also be
manually applied.

@@

Measurements of physical constants (and the units defined from them) improve
over time. Periodically the standardized systems of units and constants are
updated to reflect these improvements. For instance, in 2019 the SI system was
redefined \citep{NIST2019}, with an accompanying update to the physical
constants in
\href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
2018}}. \astropyconstants now defaults to use the \texttt{SI/CODATA 2018}
values, with the units in \astropyunits based on these constants. Most of
\astropypkg and affiliate packages build upon \astropyunits, so this update
effects the entire \astropy ecosystem.

@@

Performance in \astropypkg is addressed based on feedback and requests, and
through contributions from community members: If you find that parts of
\astropypkg are not performant, we encourage you to open a \github issue to
start a discussion with the \astropypkg core maintainers.

@@

The Astropy project supports the broader ecosystem by providing pre-configured
infrastructure packages that the community can use to support and maintain their
own software package infrastructure. These include tools and extensions that
enable easily generating documentation and setting up automated testing, as well
as providing package scaffolding for new projects.

@@

We would like to thank the members of the community who have contributed to
\astropy, who have opened issues and provided feedback, and have supported the
project in a number of different ways.
